---
title: "Sample size requirements for contact surveys"
format: html
---

```{r setup}
#| message: false
library(tidyverse)
library(boot)
library(furrr)
library(patchwork)

# Figure dimension constants (golden ratio proportions)
H <- 3
goldenratio <- (1 + sqrt(5)) / 2

# setting up parallel processing for furrr including number of cores used and 
# random seeding
no_cores <- availableCores() - 1
plan(multisession, workers = no_cores)

```

## Figure 1: Sample sizes in the contact survey literature

Reads a rapid-review dataset, filters to included studies, and plots the
distribution of sample sizes (capped at 10,000 for display).

```{r figure1}
reviewdata <- read.csv("../rapidreviewdata_update.csv")

reviewdata %>%
  filter(INCLUDE. == "YES") %>%
  group_by(Title, COUNTRY, SAMPLE.SIZE) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(ss2 = ifelse(SAMPLE.SIZE > 10000, 10000, SAMPLE.SIZE)) %>%
  ggplot(aes(x = ss2)) +
  geom_histogram(binwidth = 200, color = "darkblue", fill = "lightblue") +
  xlab("Sample size") +
  ylab("Frequency") +
  theme_bw() +
  theme(
    axis.text  = element_text(size = 14),
    axis.title = element_text(size = 20)
  ) -> f1

ggsave("Figure1.png", f1, height = H, width = H * goldenratio,
       dpi = 600, device = "png")
f1
```

## POLYMOD eigenvalue analysis

Sub-samples the POLYMOD dataset at a range of sample sizes, constructs a
contact matrix with 10-year age bands, and extracts the dominant eigenvalue
of each sampled matrix. Repeating this many times shows how stable the
dominant eigenvalue (a proxy for R0) is as a function of sample size.

**Data required:** `3874557/2008_Mossong_POLYMOD_participant_common.csv` and
`3874557/2008_Mossong_POLYMOD_contact_common.csv` (Zenodo record 3874557).

```{r polymod-setup}
participantdata <- read.csv(
  "3874557/2008_Mossong_POLYMOD_participant_common.csv"
)
contactdata <- read.csv(
  "3874557/2008_Mossong_POLYMOD_contact_common.csv"
)

maxsize  <- nrow(participantdata)   # total participants in full dataset
x        <- c(100, 500, 1000, 1500, 2000, 3000, 4000, 5000)  # sample sizes
x2       <- c(100,5800,-100)
numreps  <- 200                     # repetitions per sample size

maxeigenval <- array(NA, dim = c(length(x), numreps))
```

```{r polymod-simulation}
#| cache: true   # simulation is slow; cache to avoid re-running on re-render

for (i in seq_along(x)) {
  ss <- x[i]
  for (nr in 1:numreps) {

    # Randomly include each participant with probability ss/maxsize,
    # assign 10-year age bands (capped at 70+), join to contacts,
    # compute per-participant contact rates, and reshape to a matrix.
    participantdata %>%
      rowwise() %>%
      mutate(
        include = rbinom(1, 1, prob = ss / maxsize),
        agegp1  = ifelse(part_age <= 70, floor(part_age / 10), floor(70 / 10))
      ) %>%
      add_count(agegp1, wt = include) %>%
      right_join(contactdata %>% filter(!is.na(cnt_age_exact)),
                 by = "part_id") %>%
      filter(include == 1) %>%
      mutate(
        agegp2 = ifelse(cnt_age_exact <= 70,
                        floor(cnt_age_exact / 10), floor(70 / 10))
      ) %>%
      group_by(agegp1, agegp2, n) %>%
      summarise(totalcontacts = n(), .groups = "drop_last") %>%
      mutate(contactrate = totalcontacts / n,
             contactrate = ifelse(is.na(contactrate), 0, contactrate)) %>%
      pivot_wider(id_cols = agegp1, names_from = agegp2,
                  values_from = contactrate, values_fill = 0) %>%
      replace_na(list(Value = 0)) -> contactmatrix

    # Dominant eigenvalue of the contact matrix, excluding the oldest age group
    L1 <- ncol(contactmatrix)
    maxeigenval[i, nr] <- round(
      max(Re(eigen(contactmatrix[1:(L1 - 2), 2:(L1 - 1)])$values)), 3
    )
  }
}
```

```{r polymod-simulation-v2}
#| cache: true   # simulation is slow; cache to avoid re-running on re-render

# Extract the per-draw computation into a plain function.
# Takes a sample size ss, runs the full sub-sampling pipeline once,
# and returns the scalar dominant eigenvalue of the contact matrix.
compute_eigenval <- function(ss) {
  participantdata %>%
    rowwise() %>%
    mutate(
      include = rbinom(1, 1, prob = ss / maxsize),
      agegp1  = ifelse(part_age <= 70, floor(part_age / 10), floor(70 / 10))
    ) %>%
    add_count(agegp1, wt = include) %>%
    right_join(contactdata %>% filter(!is.na(cnt_age_exact)),
               by = "part_id") %>%
    filter(include == 1) %>%
    mutate(
      agegp2 = ifelse(cnt_age_exact <= 70,
                      floor(cnt_age_exact / 10), floor(70 / 10))
    ) %>%
    group_by(agegp1, agegp2, n) %>%
    summarise(totalcontacts = n(), .groups = "drop_last") %>%
    mutate(contactrate = totalcontacts / n,
           contactrate = ifelse(is.na(contactrate), 0, contactrate)) %>%
    pivot_wider(id_cols = agegp1, names_from = agegp2,
                values_from = contactrate, values_fill = 0) %>%
    replace_na(list(Value = 0)) -> contactmatrix

  L1 <- ncol(contactmatrix)
  round(max(Re(eigen(contactmatrix[1:(L1 - 2), 2:(L1 - 1)])$values)), 3)
}

# Build a tidy grid of every (sample size × rep) combination and map the
# function over it. Each row is one independent draw; randomness comes from
# rbinom inside compute_eigenval, so the rep column is just a counter.
maxeigenval_v2 <- crossing(ss = x2, rep = seq_len(numreps)) %>%
  mutate(maxeigen = future_map_dbl(ss, compute_eigenval,
                                   .options = furrr_options(seed = TRUE),
                                   .progress = TRUE))
```

```{r polymod-normalise}
norm_const <- mean(maxeigenval[length(x), ])
maxeigenval <- maxeigenval / norm_const
maxeigenval_v2 <- maxeigenval_v2 %>% mutate(maxeigen = maxeigen / norm_const)
```

```{r polymod-scatter}
# Scatter of normalised dominant eigenvalues across all reps, by sample size
as.data.frame(maxeigenval) %>%
  mutate(SampleSize = x) %>%
  pivot_longer(-SampleSize, names_to = "rep", values_to = "maxeigen") %>% 
    ggplot(aes(x = SampleSize, y = maxeigen)) +
    geom_point(size = 1.2, colour = "darkblue", alpha = 0.3) +
    xlab("Sample size") +
    ylab("Normalised dominant eigenvalue") +
    theme_bw() +
    theme(
      axis.text  = element_text(size = 10),
      axis.title = element_text(size = 14)
    ) -> original

maxeigenval_v2 %>%
  ggplot(aes(x = ss, y = maxeigen)) +
    geom_point(size = 1.2, colour = "darkblue", alpha = 0.3) +
    xlab("Sample size") +
    ylab("Normalised dominant eigenvalue") +
    theme_bw() +
    theme(
      axis.text  = element_text(size = 10),
      axis.title = element_text(size = 14)
    ) -> paral 

original + paral + plot_layout(axes = "collect_y") & 
  ylim(min(0.5), max(1.5))
```

```{r polymod-diffs}
# Absolute difference in eigenvalue between rep 1 and rep 2, normalised by
# the value at the largest sample size — a simple convergence diagnostic.
diffs <- abs(maxeigenval[, 1] - maxeigenval[, 2])

data.frame(SampleSize = x, normdiff = diffs / diffs[length(diffs)]) %>%
  ggplot(aes(x = SampleSize, y = normdiff)) +
  geom_point(size = 2) +
  geom_line() +
  xlab("Sample size") +
  ylab("Normalised difference (reps 1 vs 2)") +
  theme_bw() +
  theme(
    axis.text  = element_text(size = 10),
    axis.title = element_text(size = 14)
  )
```

```{r polymod-density}
as.data.frame(maxeigenval) -> mymatrix
rownames(mymatrix) <- x

# Density plot: tighter distributions at larger sample sizes indicate convergence
mymatrix %>%
  tibble::rownames_to_column(var = "ss")%>%
  mutate(ss = as.numeric(ss)) %>% 
  pivot_longer(-ss, names_to = "rep", values_to = "maxeigen") %>%
  ggplot(aes(x = maxeigen, colour = as.factor(ss))) +
    geom_density() +
    xlab("Normalised dominant eigenvalue") +
    ylab("Density") +
    labs(colour = "Sample size") +
    theme_bw()+
    coord_flip() -> density_plot




```

## Figure 2: R_t estimates across sample sizes

Uses a pre-processed contact dataset to estimate the effective reproduction
number at varying sample sizes. The weighted mean of squared contacts
(weighted by age-group size) is used as the key statistic.

**Data required:** `contactsperson_withkids6.Rdata` and `Rall.Rdata`
(pre-computed simulation output).

```{r rt-setup}
# Weighted mean helper: sum(degree^2 * weight) / sum(weight)
myweightedmean <- function(DD, ix) {
  sum(DD[ix, 1]^2 * DD[ix, 2]) / sum(DD[ix, 2])
}

# Compute per-person effective degree (number of potentially infectious contacts)
degreeperson0 <- function(contactsperson, inf_child, CTF1) {
  contactsperson %>%
    filter(dcat >= 0) %>%
    mutate(
      infect  = ifelse(age <= 11, inf_child, 1),
      Num1    = ifelse(contacttrace == TRUE,
                       ((1 - CTF1) * Numbers) * (1 - comply) * infect,
                       Numbers * (1 - comply) * infect),
      homenum  = Num1 * Home,
      worknum  = Num1 * WorkSchool,
      othernum = Num1 * OtherLeisure + Num1 * Travel
    ) %>%
    group_by(PID) %>%
    summarise(
      degree  = sum(homenum) + sum(worknum) + sum(othernum),
      totdcat = sum(dcat),
      wcon    = sum(homenum * dcat2) + sum(worknum * dcat2) +
                sum(othernum * dcat2),
      age       = mean(age),
      ageweight = mean(ageweight),
      .groups   = "drop"
    )
}
```

```{r rt-calibrate}
load("contactsperson_withkids6.Rdata")

# Set compliance and contact tracing to zero (baseline, no interventions)
contactsperson$comply       <- 0 * contactsperson$age
contactsperson$contacttrace <- rep(FALSE, nrow(contactsperson))

# Calibrate multiplier so that weighted mean degree gives meanR0 = 2.8
degall  <- degreeperson0(contactsperson, inf_child = 1, CTF1 = 0)
meanR0  <- 2.8
mult1   <- meanR0 / myweightedmean(cbind(degall$wcon, degall$ageweight), TRUE)
```

```{r rt-simulation}
#| cache: true
infchild <- 1

PIDs    <- unique(contactsperson$PID)
#x       <- c(seq(length(PIDs), 100, -100))
x       <- c(seq(5800, 100, -100))
numreps <- 200
Rall    <- matrix(0, nrow = length(x), ncol = numreps)
i <- 1
for (ss in x) {
  for (nr in 1:numreps) {
    rind <- c()
    contactsperson$use <- contactsperson$PID * 0
    mysample <- sample.int(length(PIDs), ss)
    for (s in mysample) {
      ix <- which(contactsperson$PID == mysample[s])
      contactsperson$use[ix] <- 1
    }
    degall <- degreeperson0(contactsperson[contactsperson$use == 1, ], infchild, 0.0)
    rind   <- rbind(rind, cbind(degall$wcon, degall$ageweight))
    myboot <- myweightedmean(rind)
    Rall[i, nr] <- mult1 * mean(myboot)
  }
  i <- i + 1
}
save(Rall, file = "Rall.Rdata")
```

```{r rt-load}
# Load pre-computed Rall matrix (rows = sample sizes, cols = simulation reps).
# Sample sizes run from the full dataset size down to 200 in steps of 100.

load("Rall.Rdata")
```

```{r rt-normalise}
norm_const_rt <- mean(Rall[1, ])   # largest sample size is first row
Rall <- Rall / norm_const_rt
```

```{r figure2}
as_tibble(Rall) -> Rall
Rall$SampleSize <- x

Rall %>%
  pivot_longer(cols = -SampleSize, names_to = "simno", values_to = "Rest") %>%
  ggplot(aes(x = SampleSize, y = Rest)) +
  geom_point(size = 1.8, colour = "darkblue", alpha = 0.4) +
  geom_hline(yintercept = 1,   colour = "black") +
  geom_hline(yintercept = 2,   colour = "darkgrey", linetype = "dashed") +
  geom_hline(yintercept = 0.5, colour = "darkgrey", linetype = "dashed") +
  xlab("Sample size") +
  ylab(expression("Normalised " * R[t])) +
  theme_bw() +
  theme(
    axis.text  = element_text(size = 10),
    axis.title = element_text(size = 14)
  ) -> p2

ggsave("Figure2.png", p2, height = H, width = H * goldenratio,
       dpi = 600, device = "png")
p2



```
